---
title: "Homework 2"
author: "Dima Galat"
date: "23 October 2016"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
knitr::opts_chunk$set(warning = F)
```



## Question 1
Creating histograms for quantitative variables

```{r, include = FALSE}
lib.list <- c("readxl", "dplyr", "ggplot2", "tidyr", "rpart", "rpart.plot")
lapply(lib.list, require, character.only = TRUE)
load(file.path("..", "data", "election_2012.Rda"))
```

- This is the histogram that you requested:
```{r}
state.names <- c('Indiana', 'Alaska')
df <- tab$`2012 Presidential General Election Results`
df <- df[1:535,]
df.q1 <- filter(df, STATE %in% state.names)
df.q1 <- df.q1[-c(6,19),]
AK <- df.q1[1:5,]
IN <- df.q1[6:nrow(df.q1),]

hist(IN$`GENERAL RESULTS`, freq = T, breaks = length(IN$`GENERAL RESULTS`) + 1, main = "Indiana number of results density binning", 
     xlab = "Bins by number of voters", ylab = "Frequency of values in a bin",
     xlim = c(0, max(IN$`GENERAL RESULTS`) * 1.25))
```

- These are boxplots for categorical variables that we have (party and candidate):
```{r}
pl <- ggplot(AK, aes(x = AK$`PARTY`, y = AK$`GENERAL %`))
pl + geom_bar(stat = "identity") + xlab("Party") + ylab("Proportion of voters") +
     ggtitle("Voters per party in Alaska")

pl <- ggplot(AK, aes(x = AK$`LAST NAME`, y = AK$`GENERAL %`))
pl + geom_bar(stat = "identity") + xlab("Candidate") + ylab("Proportion of voters") + 
     ggtitle("Voters per candidate in Alaska")

aggr.dat <- aggregate(IN$`GENERAL %` ~ IN$PARTY, FUN = sum)
pl <- ggplot(aggr.dat, aes(x = aggr.dat[ ,1], y = aggr.dat[ ,2]))
pl + geom_bar(stat = "identity") + xlab("Party") + ylab("Proportion of voters") +
     ggtitle("Voters per party in Indiana")

pl <- ggplot(IN, aes(x = IN$`LAST NAME`, y = IN$`GENERAL %`))
pl + geom_bar(stat = "identity") + xlab("Candidate") + ylab("Proportion of voters") +
     ggtitle("Voters per candidate in Indiana") +
     theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
```

## Question 2
Creating box and scatter plots for quantitative variables

```{r, fig.width = 12, fig.height = 10}
find.party.name.from.abbr <- function(abbr.list) {
    label.dict <- tab$`2012 Party Labels`[5:116, c(1, 3)]
    names(label.dict) <- c("abbr", "name")
    results <- c()
    for (abbr in abbr.list) {
        results <- c(results, label.dict[label.dict["abbr"] == as.character(abbr), "name"])
    }
    return(results)
}

q2 <- filter(df, is.na(`TOTAL VOTES`))
q2.aggr <- data.frame()
for (state in unique(df$STATE)) {
    d <- filter(q2, STATE == state)
    d <- aggregate(d$`GENERAL RESULTS` ~ d$PARTY, FUN = sum)
    d <- cbind(state, d)
    q2.aggr <- rbind(q2.aggr, d)
}
names(q2.aggr) <- c(names(q2.aggr)[1], "party", "votes")
q2.aggr <- mutate(q2.aggr, party = replace(party, party == "Combined Parties:", "D"))

# scatter party
ggplot(q2.aggr, aes(x = find.party.name.from.abbr(party), y = votes, color = party)) +
    geom_point(shape = 1) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1), legend.position="bottom") +
    xlab("Party") + ylab("Number of voters in a state") + ggtitle("Voters per party for all states")

# scatter candidate
names(q2)[7] <- "Candidate"
ggplot(q2, aes(x = q2$Candidate, y = q2$`GENERAL RESULTS`, color = q2$`LAST NAME`)) +
    geom_point(shape = 1) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1), legend.position="bottom") +
    xlab("Candidates") + ylab("Number of voters in a state") + ggtitle("Voters per candidate for all states")
```
    
```{r, fig.width = 10, fig.height = 10}
# boxplot party
plot(as.factor(find.party.name.from.abbr(q2.aggr$party)), 
     q2.aggr$votes, ylab = "Number of voters in a state", cex.axis = 0.5, las = 2,
     xlab = "Party", cex.lab = 1.2)

# boxplot candidate
plot(as.factor(q2$`LAST NAME`), 
     q2$`GENERAL RESULTS`, xlab= "Candidates", ylab = "Number of voters in a state", cex.axis = 0.45, las = 2)
```

## Question 3
Create an overlay based on party red and blue lines as shown in Lynda.com exercise

```{r, fig.width = 10, fig.height = 9, warning = F}
get.appendix.a.column.names <- function(df) {
    i <- 0
    names.vec <- c(tolower(df[6,1]))
    for (year in as.numeric(df[5, seq(2, ncol(df), 2)])) {
        i <- i + 2
        names.vec <- c(names.vec, 
                       paste(year, gsub("\\(|\\)", "", df[7, i]), sep = " - "),
                       paste(year, gsub("\\(|\\)", "", df[7, i + 1]), sep = " - "))
    }
    return(names.vec)
}
get.overlayed.hist <- function(dat, label, colour) {
    if (colour == "green") {
        x.lim <- c(0, max(dat) * 1.25)
    } else {
        x.lim <- c(0, 1)
    }    
    hist(dat, density = 20, breaks = 20, prob = T, 
         xlim = x.lim,
         xlab = "Proportion of people voting for a party in all states 1996 to 2012",
         main = label)
    lines(density(dat), add = TRUE, col = "brown")
    curve(dnorm(x, mean = mean(dat), sd = sd(dat)), col = colour, lwd = 2, add = TRUE, yaxt = "n")
}

df.historic <- tab$`Appendix A. Table:  1996-2012 Presidential General Election Percentage of Popular Vote Received by State (Democratic/Republican Nominees)`
names(df.historic) <- get.appendix.a.column.names(df.historic)
df.historic <- df.historic[c(8:58), ]

df.q3 <- df.historic %>% gather(df.q3, "state")
df.q3 <- cbind(df.q3, t(as.data.frame(strsplit(df.q3[,2], " - "))))
df.q3 <- df.q3[, -2]
rownames(df.q3) <- 1:nrow(df.q3)
names(df.q3) <- c(names(df.q3)[1], "value", "year", "party")

# Normalizing values
for (i in seq(1:nrow(df.q3))) {
    if (df.q3[i,"value"] >  1) {
        df.q3[i, "value"] <- as.numeric(df.q3[i, "value"]) / 100
    }
}
# Calculating missing values
for (s in unique(df.q3$state)) {
    for (y in seq(2012, 1996, -4)) {
        year.df <- filter(df.q3, year == y, state == s)
        df.q3 <- rbind(df.q3, c(s, 1 - sum(as.numeric(year.df[,"value"])), y, "O"))
    }
}

get.overlayed.hist(as.numeric(filter(df.q3, party == 'D')$value), "Democrat votes distribution", "blue")
get.overlayed.hist(as.numeric(filter(df.q3, party == 'R')$value), "Republican votes distribution", "red")
get.overlayed.hist(as.numeric(df.q3[is.na(df.q3$party),]$value), "Other votes distribution", "green") 
```

## Question 4
Calculate the frequencies of wins for each party win with all states.

```{r}
df.q4 <- filter(df, is.na(`TOTAL VOTES`), `WINNER INDICATOR` == 'W')
republican.wins <- count(filter(df.q4, `LAST NAME` == 'Romney'))[[1]]
democrat.wins <- count(filter(df.q4, `LAST NAME` == 'Obama'))[[1]]
cat(paste0("Number of republican wins: ", republican.wins, "\n"))
cat(paste0("Number of democrat wins: ", democrat.wins, "\n"))
```

## Question 5
Create a summary using descriptive statistics off voter turnout in all states. Return the values of percentage for each party.

```{r}
df.q5 <- tab$`Table 4.  2012 General Election Votes Cast by Party`
names(df.q5) <- df.q5[4,]
df.q5 <- df.q5[-c(1:4,61:nrow(df.q5)),]
df.q5$`Democratic Candidates`[is.na(df.q5$`Democratic Candidates`)] <- 0
df.q5$`Republican Candidates`[is.na(df.q5$`Republican Candidates`)] <- 0
df.q5$`Other Candidates`[is.na(df.q5$`Other Candidates`)] <- 0
sum.vec <- as.numeric(df.q5[, 2]) + as.numeric(df.q5[, 3]) + as.numeric(df.q5[, 4])
cat("Democratic turnout summary: \n")
print(summary(as.numeric(df.q5[, 2]) / sum.vec))
cat("Republican turnout summary: \n")
print(summary(as.numeric(df.q5[, 3]) / sum.vec))
cat("Other turnout summary: \n")
print(summary(as.numeric(df.q5[, 4]) / sum.vec))
```

## Question 6
Using a single proportion: Hypothesis test and confidence interval

Lynda shows an example for comparing independent samples, whereas we only
have one sample, since democrats and republican votes are not independent.
We could've used the function below, but it would've been incorrect.

```{r}
df.q6 <- filter(df, is.na(`TOTAL VOTES`), `WINNER INDICATOR` == 'W')
republican.wins <- count(filter(df.q6, `LAST NAME` == 'Romney'))[[1]]
democrat.wins <- count(filter(df.q6, `LAST NAME` == 'Obama'))[[1]]
total.voting <- republican.wins + democrat.wins

x <- republican.wins
n <- total.voting
```

```{r, echo = T, include = T}
prop.test(x, n, p = 0.5, conf.level = 0.95, alternative = "two.sided")
```
Hence, I wrote some code to do a Binomal Proportion test that makes sense here.

Since, Ho: p0=0.05 vs Ha: p0 is not 0.05, use 2 sided test; Z~Norm(0,1)
```{r}
# Binomial proportion test
smpl.prop <- x / n
p0 <- 0.5
z <- (smpl.prop - p0) / sqrt(smpl.prop * (1 - smpl.prop) / n)
alpha <- 0.05 

# 
z.norm <- qnorm(1 - alpha / 2)
if (abs(z) < z.norm) {
    cat("Accept Ho, Not enough evidence to reject Ho at 5%\n")
} else {
    cat("Statistical evidence suggests, that Ho does not hold, therefore accept Ha\n")
}
```

- Errors for extra credit

Type 1 error,  that we reject H0, given that H0 is true- will be used, if we reject NULL

Type 2 error, rejecting Ha, given that Ha is true; same as the probability that observed value is not the rejection region when Ha is true

```{r}
lower.bound <- -(z.norm * sqrt(smpl.prop * (1 - smpl.prop) / n)) + p0
upper.bound <- (z.norm * sqrt(smpl.prop * (1 - smpl.prop) / n)) + p0

republican.wins <- filter(df.q6, `LAST NAME` == 'Romney')
pHa <-  round(mean(republican.wins$`GENERAL %`), 1) 

rej.reg.upper.bound <- (upper.bound -pHa) * sqrt(n) / sqrt(smpl.prop * (1 - smpl.prop))
rej.reg.lower.bound <- (lower.bound - pHa) * sqrt(n) / sqrt(smpl.prop * (1 - smpl.prop))
beta <- 1 - pnorm(rej.reg.upper.bound) + pnorm(rej.reg.lower.bound)
cat(paste0("The value of the alpha is ", round(alpha * 100, 1) , "%", "\n"))
cat(paste0("The value of the beta is ", round(beta * 100, 1) , "%", "\n"))
```

## Question 7
Using a single mean: Hypothesis test and confidence interval

```{r}
df.q7 <- read_excel(file.path("..", "data", "US Presidential Results & PVIs by State 1828-2012.xlsx"))
df.q7 <- df.q7[df.q7[1] == "Indiana", ]
names(df.q7) <- as.numeric(names(df.q7))
hist.subset <- data.frame()
for (year in seq(2012, 1960, -4)) {
    col.num <- grep(year, colnames(df.q7))[1]
    data.row <- c(as.numeric(year), as.numeric(df.q7[,c(col.num, col.num + 1)]))
    hist.subset <- rbind(hist.subset, data.row)
}
names(hist.subset) <- c("year", "democrat", "republican")
```

Claim: Democrates will lose the election with mu < 0.5

Null Hypothesis is that Republicans will score 50% of voters, therefore win

Alternative hypothesis: Republicans will get less than 50 % of voters and lose

Test to perform at alpha 5% significance:

    Ho: mu = 0.5001
    
    Ha: mu < 0.5001

```{r}
muH0 <-  0.5001 
alpha <- 0.95

# sample mean and sample standard deviation
sample.mu <- mean(hist.subset[, 3])
sample.s <- sd(hist.subset[, 3])
# calculation of the test statistics under Ho; mu=0.5001
ts <- (sample.mu - muH0) / sample.s * sqrt(nrow(hist.subset))
# perform a one sided t-test. Rejection region RR={ts<t value at 95%}, dof of t.test are (number of sampled values munis 1)
qt(1 - alpha, nrow(hist.subset))
if (ts > qt(1 - alpha, nrow(hist.subset))) {
    cat("Statistical evidence suggests that Ho cannot be rejected at 5%, because the sample does not significantly deviate from mu\n") 
} else {
    cat("There is enough evidence to reject Ho at alpha=5 significance level, because sample significantly deviate from mu=0.5\n")
}

# # Alternatively, we can test using p value
# ttest <- t.test(x = (hist.subset[, 2]), mu = muH0, alternative="less", conf.level = alpha)
# if (ttest$p.value > 0.05) {
#     print("Statistical evidence suggests that Ho cannot be rejected at 5%, because the sample does not significantly deviates from mu=0.5") 
# } else {
#     print("There is enough evidence to reject Ho, because sample mean significantly deviate from mu=0.5")
# }
```



## Question 8
Using a single categorical variable: One sample chi-square test

This is a goodness of fit test, we are testing that votes are equally distributed between parties, this is an equivalent of saying mu = 0.5
    Ho: votes are equally distributed between both parties
    Ha: votes are not equally distributed between both parties
    
```{r}
voters <- colMeans(hist.subset[,2:3]) * 100

chi1 <- chisq.test(voters)
chi1

if (chi1$p.value > 0.05) {
    cat("Statistical evidence suggests that Ho cannot be rejected at 5%, because the sample does not significantly deviate from mu=0.5\n") 
} else {
    cat("There is enough evidence to reject Ho, because sample significantly deviate from mu = 0.5\n")
}
```

## Question 9
Create a decision tree in your state based on voter turnout. 
```{r}
IN <- filter(df, STATE == state.names[1], is.na(`TOTAL VOTES`))
IN <- aggregate(IN$`GENERAL %` ~ IN$PARTY, FUN = sum)
names(IN) <- c("party", "result")
fit <- rpart(result~., method = "anova", data = IN, 
             control = rpart.control(minsplit = 2, minbucket = 1))
prp(fit, main = "Regression Tree based on Party voting turnout in Indiana")

```


## Question 10
Create a scatter plot with liner regression line to show each counties voting pattern for each party.

```{r}
df.historic <- tab$`Appendix A. Table:  1996-2012 Presidential General Election Percentage of Popular Vote Received by State (Democratic/Republican Nominees)`
names(df.historic) <- get.appendix.a.column.names(df.historic)
df.historic <- df.historic[c(8:58), ]

df.q10 <- filter(df.historic, state == state.names[1]) %>% gather(df.q10, "state")
df.q10 <- cbind(df.q10, t(as.data.frame(strsplit(df.q10[,2], " - "))))
df.q10 <- df.q10[, -2]
rownames(df.q10) <- 1:nrow(df.q10)
names(df.q10) <- c(names(df.q10)[1], "value", "year", "party")

# Normalizing values
for (i in seq(1:nrow(df.q10))) {
    if (df.q10[i,"value"] >  1) {
        df.q10[i, "value"] <- as.numeric(df.q10[i, "value"]) / 100
    }
}
# Calculating missing values
for (s in unique(df.q10$state)) {
    for (y in seq(2012, 1996, -4)) {
        year.df <- filter(df.q10, year == y, state == s)
        df.q10 <- rbind(df.q10, c(s, 1 - sum(as.numeric(year.df[,"value"])), y, "O"))
    }
}

df.q10$colour[df.q10$party == "R"] = "red"
df.q10$colour[df.q10$party == "D"] = "blue"
df.q10$colour[is.na(df.q10$party)] = "green"
plot(df.q10$year, df.q10$value, col = df.q10$colour, ylim=c(0, max(as.numeric(df.q10$value)) * 2),
     main = "Indiana voting patterns", xlab = "Year", ylab = "Proportion of voters")

legend('topright',  c("D", "R", "O"), 
       lty = 1, col=c("blue", "red", "green"), bty = 'n', cex = .5)
abline(lm(as.numeric(df.q10$value[df.q10$party == "D"]) ~ as.numeric(df.q10$year[df.q10$party == "D"])), col = "blue")
abline(lm(as.numeric(df.q10$value[df.q10$party == "R"]) ~ as.numeric(df.q10$year[df.q10$party == "R"])), col = "red")
abline(lm(as.numeric(df.q10$value[is.na(df.q10$party)]) ~ as.numeric(df.q10$year[is.na(df.q10$party)])), col = "green")
```