---
title: "Homework 2"
author: "Dima Galat"
date: "3 October 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{retrieving data}
data.url <- "https://canvas.harvard.edu/courses/18032/files/2908415/download?wrap=1"
data <- read_exel(data.url)
```

```{r cars}
summary(cars)
```

## Homework questions
1) Creating histograms for quantitative variables
Create a histogram of your assign STATE for our group project. You will plot the bins based on party and candidates.
?! "Table 2. Electoral & Pop Vote" C1 = STATE
"2012 Pres General Results"
"2012 Pres Primary Results"

2) Creating box plots for quantitative variables
Create a box and scattered plot based on party and candidates.

3) Overlaying plots
Create an overlay based on party red and blue lines as shown in Lynda.com exercise.

4) Calculating frequencies
Calculate the frequencies of wins for each party win with all states.

5) Calculating descriptive
Create a summary using descriptive statistics off voter turnout in all states. Return the values of percentage for each party.

6) Using a single proportion: Hypothesis test and confidence interval
Create a hypothesis statement based on your state regarding who will win the 2016 President Election. Use the historical l data provided the above mention data set.

7) Using a single mean: Hypothesis test and confidence interval
Create a Hypothesis test in RStudio based on the historical data regarding the outcome of your state. Find a source that shows the results from 1960 to 2012 and state the probability of the historical data who will win in 2016.  Use your state only!
You must state both the Null (Ho) and Alternative (Ha).

8) Using a single categorical variable: One sample chi-square test
State how you would create a one sample Pearson chi-square test to test your hypothesis in question 7. Create a sample in RStudio based on Lynda.com practice exercise.

9) Create a decision tree in your state based on voter turnout. Group your candidates in the following category based off of the historical data provided by the FEC.
Clinton, Johnson, Stine, and Trump are the candidates. Based off of percentages off voter turnout in 2012 who will win your state.  So you will put them into categories like D (Dem), R (Rep), and O (Other).

10) Create a scatter plot with liner regression line to show each counties voting pattern for each party
Select your state and their counties to plot the strengthen and weakness for each district. Use a color scheme that match the party colors – red, blue, and green.  Yes, green is for other.


## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
